Kepler's Demographic Cliff: Ages of Young Stars and Planets in the Kepler Field

### 2024-05-20 paper/ms.tex
update ms

% <250 words
\begin{abstract}
  %
  Recent analyses of FGK stars in open clusters have helped clarify
  the precision with which a star's rotation rate and lithium content
  can be used as empirical indicators for its age.
  %
  Here we apply this knowledge to stars observed by Kepler.
  %
  Rotation periods are drawn from previous Kepler analyses; lithium is
  measured from new and archival Keck/HIRES spectra.
  %
  We report rotation-based ages for \nuniqstarsantosrotgyroappl\ stars
  and \nplwgyroagenograzing\ planets for which our gyrochronology
  model is applicable; these rotation-based ages are over an order of
  magnitude more precise than isochrone ages in this sample of stars.
  %
  The independent lithium ages agree with the rotation-based ages in
  $\approx$90\% of cases, suggesting that most of the rotation-based
  ages are not only precise, they are also accurate.
  %
  Our resulting catalog includes \nplyounggyrotwosigma\ planets
  younger than 1\,Gyr at 2$\sigma$ (\nplyounggyrothreesigma\ planets
  at 3$\sigma$).
  %
  More broadly, we find that the star formation rate in the Kepler field
  has dropped by a factor of \ratiosfr$\pm$\uncratiosfr\ over the past
  3,Gyr. This trend holds true for both planet hosts and the overall
  stellar sample.
  %
  This ``demographic cliff'' is a known feature of the Galaxy's star
  formation history, and it helps clarify the relative balance of
  young and old planets that have been detected to date using
  transits.
  %
\end{abstract}

@@

Exoplanet science is enjoying an age of abundance, fueled by the discovery
of thousands of worlds orbiting close to their host stars
\citep{Borucki10,2015JATIS...1a4003R}.  Most of these exoplanets are
billions of years old.  This fact leaves many gaps in our knowledge of
the exact physical and dynamical pathways that produce these objects.
The reason is that processes such as thermal cooling
\citep{2007ApJ...659.1661F}, atmospheric loss
\citep{2019AREPS..47...67O}, giant impacts
\citep{2014prpl.conf..595R}, and dynamical instabilities
\citep{2017MNRAS.470.1750I} are expected to be most efficient over
$\approx$10\,Myr to $\approx$1\,Gyr timescales. For most known
exoplanets, these processes have already run their course.

Young ($<$1\,Gyr) exoplanets represent one means of building the timeline for
exoplanet evolution.  Informative individual young systems include
V1298~Tau, an adolescent resonant chain of four transiting puffy
planets that is a likely precusor to the compact multiplanet systems
\citep{David_2019}, and HIP~67522, a Jupiter-sized planet with a
Neptunian mass (\citealt{Rizzuto_2020}; Thao et al.~submitted).
Population-level analyses have similarly suggested differences in the
size distribution of young exoplanets relative to their older
counterparts
\citep{Berger_2020b_rpage,David_2021,Sandoval_2021,2023AJ....166..248C,2024arXiv240303261V}.

Discovering a young planet requires solving two problems: find the
planet, and measure the star's age.  Each problem admits a range of
solutions \citep[e.g.][]{2008Sci...322.1348M,2012ApJ...756L..33Q,2024AJ....167..193T}.  In
this article we will consider planets previously known from transits,
and infer new stellar ages using rotation and lithium.

To begin, imagine that the ages of nearby stars in the Galaxy are
uniformly distributed from 0--10\,Gyr \citep[e.g.~Fig~17
of][]{Nordstrom_2004}.  This approximation clarifies the rarity of
young stars: under this assumption, $\approx$1\% of nearby stars have
$t$$<$100\,Myr, and $\approx$10\% have $t$$<$1\,Gyr.  Studies of
currently forming protoplanets \citep{2018A&A...617A..44K}, and of
exoplanets evolving just after disk dispersal
\citep[e.g.][]{2022MNRAS.512.5067K}, are thereby capped in their
maximum achievable sample sizes by the tyranny of the galactic star
formation rate.

Despite this and other observational challenges, young close-in planet
discovery has matured over the past decade, primarily due to Kepler,
K2, TESS, and Gaia
\citep[e.g.][]{Meibom_2013,Mann_K2_25_2016,Curtis_2018,Livingston_2018,David_2019,Bouma_2020_toi837,Rizzuto_2020,Plavchan_2020,Newton_2021,Nardiello_2022,Barber_2022,Bouma_2022b,Zhou_2022,Zakhozhay_2022,Wood_2023}.
The strategy pursued by most groups during the 2010s was to focus on
stars with known ages---obvious members of open clusters---and to
search them for transiting planets.  The resulting stellar (and
assumed planetary) ages are precise at the $\approx$10\% level.  A
more recent development, facilitated by the second Gaia data release,
is the idea that finding transiting planets is often easier than
finding the host stellar association \citep[e.g.][]{Tofflemire_2021}.

The main problem with the ``cluster-first'' approach is that most
nearby sub-gigayear stars are in the field.  In other words,
$\lesssim$1\% of stars within the nearest $\approx$500\,pc have been
associated with their birth cluster
\citep[e.g.][]{Zari_2018,CantatGaudin_2020,Kounkel_2020,Kerr_2021}.
One can appreciate the problem by querying NASA's Exoplanet Archive
\citep[NEA;][]{2013PASP..125..989A} for transiting planets younger
than 1\,Gyr.  Imposing reasonable precision requirements
(e.g.~$t$$<$1\,Gyr at 2$\sigma$, or else ages precise to say S/N$=$3)
gives $\approx$50 ``young'' transiting planets at the time of writing.
Most of these planets are in clusters, and were found by Kepler, K2,
or TESS.  However, combined, those surveys have discovered
$\approx$5{,}000 planets.  
Barring a major discovery bias against sub-gigayear planets,
a constant star formation history would imply that an order of magnitude
more young transiting planets should be known.

This study aims to resolve two main questions.  First, how wrong is it
to assume a uniform age distribution for transiting planet host stars?
Second, where are the missing young planets?  Briefly, we will find
that uniform is wrong by a factor of a few, and that increased stellar
activity for young stars may be a less important bias than the
challenge of determining precise ages for field stars.

%FIXME these paragraphs need improvements in how they incorporate
%Lu+21 (meh), but certainly Mathur+2023.

Given that Kepler's main mission ended in 2013, one might be forgiven
for assuming that the ages of the Kepler planets are already known.
Of course, they have been studied: \citet{Berger_2020b_rpage} and
\citet{Petigura_2022} for instance calculated isochrone ages for
Kepler Objects of Interest (KOIs).  These ages are most precise for
stars whose luminosities and temperatures separate them from the
main-sequence.  \citet{David_2021} leveraged isochronal ages from an
earlier analogous study \citep{Fulton_2018}, and also used a relative
gyrochrone age-ranking approach.  These studies were conducted in the
broader context of investigating trends in exoplanet demographics,
which also depend on stellar metallicity and mass
\citep[e.g.][]{Petigura_2018}.  These studies offered limited
sensitivity to the ages of $\lesssim$1\,Gyr planets, due to the nature
of isochronal dating.

Gyrochronology offers ages precise to $\lesssim$30\% for FGK stars
between 1-4\,Gyr, and provides more complex asymmetric posteriors at
earlier times \citep{Bouma_2023}.  The idea of using a star's
spin-down as a clock has a rich history
\citep{Skumanich_1972,Noyes_1984,Kawaler_1989,Barnes03,Mamajek_2008,Angus_2015},
and physics-based models for the spin-down itself can clarify many
aspects of how the stellar winds, internal structure, and magnetic
dynamo all evolve
\citep[e.g.][]{Matt_2015,Gallet_Bouvier_2015,Spada_2020}.  In the
context of Kepler, empirical models have been previously applied to
derive rotation-based ages for subsets of the Kepler data
\citep{Walkowicz_2013,Reinhold_2015,Angus_2018}.  ``Gyro-kinematic''
ages have also been reported \citep{2021AJ....161..189L}.  There have
been major improvements since these pioneering studies in:
\begin{enumerate}[label={\it \roman*)},leftmargin=*,topsep=0pt,itemsep=-1ex,partopsep=1ex,parsep=1ex]
  \item The extent of the Kepler Object of Interest catalog, and our
    vetting of false positives within it;
  %
  \item Available measurements of stellar rotation periods in open
    clusters
    \citep[e.g.][]{Curtis_2019_ngc6811,Gillen_2020,Rampalli_2021,Fritzewski_2021,Rebull_2022,Dungee_2022,2023AJ....166...14B};
  %
  \item Our ability to model not only the mean evolution but also the
    intrinsic dispersion in rotation through time \citep{Bouma_2023}.
    For Sun-like stars, the new model is broadly consistent with older
    calibrations \citep[e.g.][]{Mamajek_2008}.  For late G dwarfs and
    K dwarfs, the new calibration differs by factors of a few;
  %
  \item Our ability to identify binary stars, which can be problematic
    for estimates of rotation-based ages.
\end{enumerate}
These improvements form the basis for why we believe the present
analysis to be necessary.

Beyond rotation-based ages, recent years have also yielded
improvements in the lithium age scale.  Lithium ages include two
distinct regimes: depletion boundary ages for M dwarfs and brown dwarfs in star
clusters, and the less precise decline-based ages for individual field
FGK stars \citep{Soderblom_2010}.  Here we focus on the latter
approach, which relies on the observation that the Li abundances of
partially-convective stars gradually decline as they age
\citep[e.g.][]{2005A&A...442..615S}.  The theoretical explanation for
this decline is debated
\citep[e.g.][]{1995ApJ...441..865C,2010ApJ...716.1269D,2019MNRAS.485.4052C}.
Empirical understanding has however improved due to the work by
\citet{Jeffries_2023}, who modeled the evolution of 
the \ion{Li}{1} 6708\,\AA\ doublet equivalent width (EW) through
time using a set of 6{,}200 stars in 52
open clusters.  Two-sided lithium ages are useful for Kepler (FGK)
stars between $\approx$0.03-0.5\,Gyr, with a strong dependence on
spectral type since the K dwarfs lose their surface lithium much
faster than G dwarfs.  The precision of the ages reported by the
\citet{Jeffries_2023} method in this regime are in the range of
0.3-0.5~dex.

We discuss our method for selecting the star and planet samples in
Section~\ref{sec:selection}, and describe the origin of our adopted
stellar parameters other than ages in Section~\ref{sec:stellarprops}.
We describe and validate our age-dating methods in
Sections~\ref{sec:rotage} and~\ref{sec:liage}.  The population-level
trends for both stars and planets are described in
Sections~\ref{sec:results} and~\ref{sec:disc}, and a few conclusions
are offered in Section~\ref{sec:conclusions}.

@@

{\it Near the main-sequence (bit 9)}---Many stars observed by Kepler
are far from the main sequence: this can be seen in both panels of
Figure~\ref{fig:stellarprops}.  This is a challenge for rotation-based
ages because {\it i)} point sources far from the main-sequence include
unresolved photometric binaries, and {\it ii}) effects other than
magnetic braking influence the rotation periods of bona fide single
stars far from the main-sequence.  The added complications of
reddening and metallicity further complicate matters.  After exploring
various options, we settled on the orange locus in the $\log
g$--$T_{\rm eff}$ plane shown in Figure~\ref{fig:stellarprops}.  While
the details of how exactly this locus is constructed are arbitrary, 
%and convoluted,\footnote{In practice, for K dwarfs we fitted an
%$N^{\rm th}$-order polynomial to the set of stars for which $t_{\rm
%B20,iso}/(+\sigma_{t,{\rm B20,iso}})\approx 3$, where
%$t_{\rm B20,iso}$ was the isochronal age reported by
%\citet{Berger_2020a_catalog}. We let the order of the fit vary from
%$N$=1 to 10, minimized the Bayesian Information Criterion,
%and found $N$=6.  For F and G dwarfs, this procedure failed to produce
%a reasonable locus because of shifts in the relative precision of
%the isochrone ages.  We therefore visually selected KOIs that appeared
%to be near the ``main sequence'', fitted a separate polynomial to
%their locations, and then stitched the resulting polynomials from the
%F and G dwarfs to that of the K dwarfs.}
the general aim is to flag stars for which there is evidence based on
their location in the $\log g$--$T_{\rm eff}$ plane that our
gyrochronology model may not be reliable.  
%The rotation period
%distribution of sources excluded in this manner includes a relatively
%larger fraction of ``rapid rotators'' than the main sequence sources,
%likely due to a combination of astrophysical and systematic effects.
See for instance Figure~8a of \citet{2022AJ....164..137K} or
Figures~12 and 13 of \citet{2023ApJS..268....4F} for examples of how
the period distribution of this population differs from stars near the
main sequence.

@@

In part because many of our EWs are upper limits, many of the lithium
ages are lower limits.  In such cases,  the inferred age posterior is
strongly influenced by {\it i)} the assumed intrinsic dispersion in
the lithium model, and {\it ii)} by the prior.  In these instances, we
quote the 3$\sigma$ (99.7$^{\rm th}$ percentile) limits.

@@

{\it Occurrence Rates}:
The age-dependent trends predicted for exoplanet populations, such as
the Kelvin-Helmholtz cooling of mini-Neptunes \citep{Gupta_2019},
time-dependent carving of the photoevaporation desert
\citep{Owen_Lai_2018}, and the time evolution of the radius valley
\citep{Rogers_2021} can be explored using our data.  We defer this
analysis to a separate publication.

@@

\startlongtable
\begin{deluxetable*}{llllllllrrrrrrl}
  \tabletypesize{\scriptsize}
  \tablecaption{Ages of Kepler planets and planet candidates.  This
  version of the table is truncated to include only the youngest
  confirmed Kepler planets, sorted by the minimum of either the
  rotation or lithium-based age.  The full machine-readable table
  contains ages and age limits for \nnonfopkoissomeageinfo\ non-false
  positive KOIs with MES$>$10.  
  A bash script to decode any given $Q_{\rm star}$ quality flag is
  {\bf
  \href{https://gist.github.com/lgbouma/20368253f1a98da1b39cf32fdda0be13}{available
  online}}.  A python script to select stars with specific bit flags
  is {\bf \href{https://gist.github.com/lgbouma/87ad8bde42625e766a5a8857cc5a183a}{also available}}.
  \label{tab:planets}}
  %\toprule
  %\midrule
  %\endhead
  \startdata
  KOI & Kepler &  $T_{\rm eff}$ & $P_{\rm rot}$ & EW$_{\rm Li}$ &
  $t_{\rm gyro}$ & $t_{\rm Li} $ & Consistent? & $R_{\rm p}$ & $P_{\rm orb}$ &
  $Q_{\rm planet}$ & $Q_{\rm star}$ &  Spec? & Comment \\
  -- &   -- & K & days & m\AA & Myr & Myr &  str &   Earths &    days &       int  & int & bool & -- \\
  \hline
  \input{table_subgyr_confirmed} %\\
% \rule
%  \input{table_subgyr_candidate}
  \enddata
  \tablecomments{Two checkmarks denotes systems with two-sided $t_{\rm
  gyro}$ and $t_{\rm Li}$ for which no important quality flags are
  raised.  One checkmark denotes two-sided $t_{\rm gyro}$ only. 
  ``Confirmed'' planets appear in the machine-readable
  version before``candidate'' planets.  The bit quality flags for the
  rotation-based ages, $Q_{\rm star}$, are described in
  Section~\ref{subsec:flags}.  Concisely summarized, they are:
  Bit 0: $T_{\rm eff}/{\rm K} \in [ 3800-6200]$ ?
  Bit 1: $\log g$$<$4.2?
  Bit 2: $M_{\rm G}$<3.9 or $M_{\rm G}>8.5$?
  Bit 3: In KEBC?
  Bit 4: Large $d_{\rm xm,Kep-Gaia}$?
  Bit 5: Confused Kep-Gaia crossmatch?
  Bit 6: Gaia DR3 non-single star?
  Bit 7: RUWE$>$1.4?
  Bit 8: Crowded?
  Bit 9: Far from main-sequence?
  Bit 10: \citetalias{Santos_2021} CP/CB?
  Bit 11: $P_{\rm rot}$ not in the homogeneous
  \citetalias{Santos_2019} \citetalias{Santos_2021} sample?
  As an example, Kepler-1627 has $Q_{\rm star}$ flagged with bit 10 and bit 7.
  The analogous planet quality bitmask, $Q_{\rm planet}$, has the
  following meaning.
  Bit 0: Candidate not reliable?
  Bit 1: MES$<$10?
  Bit 2: Grazing?
  For a star to have a high likelihood of being ``reliable for
  gyrochronology'' we suggest $Q_{\rm star}$ bits zero to nine to
  not be raised, and for a planet to be ``reliabe'', we suggested
  $Q_{\rm planet}$ to be zero.
  }
\end{deluxetable*}

>>>>

% <250 words
\begin{abstract}
  %
  Recent analyses of FGK stars in open clusters have helped clarify
  the precision with which a star's rotation rate and lithium content
  can be used as empirical indicators for its age.
  %
  Here we apply this knowledge to stars observed by Kepler.
  %
  Rotation periods are drawn from previous work; lithium is
  measured from new and archival Keck/HIRES spectra.
  %
  We report rotation-based ages for \nuniqstarsantosrotgyroappl\ stars
  and \nplwgyroagenograzing\ planets for which our method is
  applicable.
  %
  We find that our rotation-based ages accurately recover ages of
  known Kepler field clusters; they also agree with
  $\gtrsim$90\% of the independent lithium ages.
  %
  Our resulting catalog includes \nplyounggyrotwosigma\ planets
  younger than 1\,Gyr at 2$\sigma$ (\nplyounggyrothreesigma\ planets
  at 3$\sigma$).
  %
  More broadly, we find that the star formation rate in the Kepler field
  has dropped by a factor of \ratiosfr$\pm$\uncratiosfr\ over the past
  3\,Gyr.  This trend holds true for both planet hosts and the overall
  stellar sample.
  %
  This ``demographic cliff'' is a known feature of the Galaxy's star
  formation history, and it helps clarify the relative balance of
  young and old planets that have been detected to date using
  transits.
  %
\end{abstract}

@@

Exoplanet science is enjoying an age of abundance, fueled by the
discovery of thousands of worlds orbiting close to their host stars
\citep{Borucki10,2015JATIS...1a4003R}.  Most of these exoplanets are
billions of years old.  This fact leaves many gaps in our knowledge of
the exact physical and dynamical histories of these objects.  The
reason is that processes such as thermal cooling
\citep{2007ApJ...659.1661F}, atmospheric loss
\citep{2019AREPS..47...67O}, giant impacts
\citep{2014prpl.conf..595R}, and dynamical instabilities
\citep{2017MNRAS.470.1750I} are expected to be most efficient over
timescales of much less than 1\,Gyr.  For most known exoplanets, these
processes have run their course.

Young ($<$1\,Gyr) exoplanets therefore represent one means of building
the timeline for exoplanet evolution.  Informative individual
exemplars include V1298~Tau, an adolescent resonant chain of four
transiting puffy planets that is a likely precusor to the compact
multiplanet systems \citep{David_2019}, and HIP~67522, a Jupiter-sized
planet with a Neptunian mass (\citealt{Rizzuto_2020}; Thao et
al.~submitted).  Population-level analyses have similarly suggested
differences in the size distribution of young exoplanets relative to
their older counterparts
\citep{Berger_2020b_rpage,David_2021,Sandoval_2021,2023AJ....166..248C,2024arXiv240303261V}.

Discovering a young planet requires solving two problems: find the
planet, and measure the star's age.  Each problem admits a range of
solutions
\citep[e.g.][]{2008Sci...322.1348M,2012ApJ...756L..33Q,2024AJ....167..193T}.
In this article we will consider planets whose existence is previously
known from transits, and infer new stellar ages using rotation and
lithium.

To begin, imagine that the ages of nearby stars in the Galaxy are
uniformly distributed from 0--10\,Gyr
\citep[][]{2000MNRAS.318..658B,Nordstrom_2004}.  This approximation
clarifies the rarity of young stars: under this assumption,
$\approx$1\% of nearby stars have $t$$<$100\,Myr, and $\approx$10\%
have $t$$<$1\,Gyr.  Studies of currently forming protoplanets
\citep{2018A&A...617A..44K}, and of exoplanets evolving just after
disk dispersal \citep[e.g.][]{2022MNRAS.512.5067K}, are thereby capped
in their maximum achievable sample sizes by the tyranny of the
galactic star formation rate.

Despite this and other observational challenges, young close-in planet
discovery has matured over the past decade, primarily due to Kepler,
K2, TESS, and Gaia
\citep[e.g.][]{Meibom_2013,Mann_K2_25_2016,Curtis_2018,Livingston_2018,David_2019,Bouma_2020_toi837,Rizzuto_2020,Plavchan_2020,Newton_2021,Nardiello_2022,Barber_2022,Zhou_2022,Zakhozhay_2022,Wood_2023}.
The strategy pursued by most groups during the 2010s was to focus on
stars with known ages---obvious members of open clusters---and to
search them for transiting planets.  The resulting stellar (and
assumed planetary) ages are precise at the $\approx$10\% level.  A
more recent development, facilitated by the second Gaia data release,
is the idea that finding transiting planets is often easier than
finding the host stellar association \citep[e.g.][]{Tofflemire_2021}.

The ``cluster-first'' approach has a major limitation: most nearby
sub-gigayear stars are in the field.  Only $\lesssim$1\% of stars
within $\approx$500\,pc have been associated with their birth cluster
\citep[e.g.][]{Zari_2018,CantatGaudin_2020,Kounkel_2020,Kerr_2021}.
One can appreciate the problem by querying the NASA Exoplanet Archive
\citep[NEA;][]{2013PASP..125..989A} for transiting planets younger
than 1\,Gyr.  Imposing a precision requirement of say $t$$<$1\,Gyr at
2$\sigma$ precision gives $\approx$50 ``young'' transiting planets at
the time of writing.  Most of these young planets are in clusters, and
were found by Kepler, K2, or TESS.  However, these surveys have
discovered $\approx$5{,}000 planets in total.  Assuming a constant
star formation history, we would expect an order of magnitude more
sub-gigayear transiting planets.

This study aims to resolve two main questions.  First, how wrong is it
to assume a uniform age distribution for transiting planet host stars?
Second, where are the missing young planets?  We will find that
uniform is wrong by a factor of a few, and that stellar activity may
be a less significant bias than the challenge of precisely determining
ages for field stars.

Despite Kepler's main mission ending quite some time ago, the ages of
many Kepler planets remain uncertain.  While isochrone ages have been
calculated for Kepler stars \citep{Berger_2020b_rpage} and Kepler
Objects of Interest \citep[KOIs;][]{Petigura_2022}, isochrone ages are
most precise for stars whose luminosities and temperatures separate
them from the main sequence.  For $\lesssim$1\,Gyr stars, isochrone
ages are therefore limited.

Stellar rotation periods offer a promising alternative.  The idea of
using a star's spin-down as a clock has a rich history
\citep{Skumanich_1972,Noyes_1984,Kawaler_1989,Barnes03,Mamajek_2008,Angus_2015}.
Modern models yield ages precise to $\lesssim$30\% for FGK stars
between 1-4\,Gyr, and constraining age posteriors at earlier times
\citep{Bouma_2023}.  Physics-based models
\citep{Matt_2015,Gallet_Bouvier_2015,Spada_2020} are beginning to
connect these empirical relations to the evolution of stellar winds,
internal structure, and angular momentum transport.

In the context of Kepler, rotation-based ages have been reported for
planet-hosting stars since the early data releases
\citep{Walkowicz_2013,Reinhold_2015,Angus_2018}.  Recent studies
focused on rotation-based ages for Kepler have further explored
incorporating information from stellar kinematics
\citep{2021AJ....161..189L,2024AJ....167..159L}, and from light curve
features such as starspot amplitudes \citep{2023ApJ...952..131M}.  Our
analysis is motivated by a few factors that we believe constitute
improvements, particularly for the ages of sub-gigayear stars.   These
factors are as follows.
\begin{enumerate}[label={\it \roman*)},leftmargin=*,topsep=0pt,itemsep=-1ex,partopsep=1ex,parsep=1ex]
  \item The Kepler Object of Interest catalog, and our vetting of
    false positives within it, has reached maturity
    \citep[e.g.][]{Thompson_2018};
  %
  \item Measured rotation periods of FGK stars in open clusters now
    show not only the average evolution of $P_{\rm rot}(T_{\rm eff},t)$,
    but also how the astrophysical dispersion of stars around this average
    converges by the $\approx$700\,Myr age of Praesepe
    \citep[e.g.][]{Curtis_2019_ngc6811,Gillen_2020,Rampalli_2021,Fritzewski_2021,Rebull_2022,Dungee_2022,2023AJ....166...14B};
  %
  \item We can fit the open cluster data, and use these fits to infer
    ages by marginalizing over the range of ages that might explain
    any one star's rotation period \citep{Bouma_2023}.
    For Sun-like stars, this model is consistent with older
    calibrations \citep[e.g.][]{Mamajek_2008}, though with an improved treatment
    of uncertainty propagation.  For late G dwarfs and
    K dwarfs, the calibration differs by factors of a few;
  %
  \item We can identify binary stars at greater fidelity today than
    in the past, which can clarify otherwise problematic estimates of
    rotation-based ages.
  %
  \item New open clusters in various stages of dissolution have been
    found in the Kepler field
    \cite[e.g.][]{2019AJ....158..122K,Bouma_2022b,2022AJ....164...88B}.
    These clusters offer a sample of stars for which we can test the
    reliability of single-star age-dating methods.
\end{enumerate}

Independently, recent years have also yielded improvements in the
lithium age scale.  Lithium ages include two distinct regimes:
depletion boundary ages for M dwarfs and brown dwarfs in star
clusters, and the less precise decline-based ages for individual field
FGK stars \citep{Soderblom_2010}.  Here we focus on the latter
approach, which relies on the observation that the Li abundances of
partially-convective stars gradually decline as they age
\citep[e.g.][]{2005A&A...442..615S}.  The theoretical explanation for
this decline is debated
\citep[e.g.][]{1995ApJ...441..865C,2010ApJ...716.1269D,2019MNRAS.485.4052C}.
Empirical understanding has however improved due to the work by
\citet{Jeffries_2023}, who modeled the time evolution of the
\ion{Li}{1} 6708\,\AA\ equivalent width (EW) using a set of 6{,}200
stars in 52 open clusters.  Two-sided lithium ages are useful for
Kepler (FGK) stars between $\approx$0.03-0.5\,Gyr, with a strong
dependence on spectral type since the K dwarfs lose their surface
lithium much faster than G dwarfs.  The precision of the ages reported
by the \citet{Jeffries_2023} method in this regime are in the range of
0.3-0.5~dex.

We discuss our method for selecting the star and planet samples in
Section~\ref{sec:selection}, and describe the origin of our adopted
stellar parameters other than ages in Section~\ref{sec:stellarprops}.
We describe our age-dating methods in
Sections~\ref{sec:rotage} and~\ref{sec:liage}, and test them using
stars in clusters in the Kepler field.  The population-level
trends for both stars and planets are described in
Sections~\ref{sec:results} and~\ref{sec:disc}, and a few conclusions
are offered in Section~\ref{sec:conclusions}.

@@

{\it Near the main sequence (bit 9)}---Many stars observed by Kepler
are far from the main sequence: this can be seen in both panels of
Figure~\ref{fig:stellarprops}.  This is a challenge for rotation-based
ages because {\it i)} point sources far from the main sequence include
unresolved photometric binaries, and {\it ii}) effects other than
magnetic braking influence the rotation periods of bona fide single
stars far from the main sequence.  The added complications of
reddening and metallicity further complicate matters.  After exploring
various options, we settled on the orange locus in the $\log
g$--$T_{\rm eff}$ plane shown in Figure~\ref{fig:stellarprops}.  While
the details of how exactly this locus is constructed are arbitrary, 
%and convoluted,\footnote{In practice, for K dwarfs we fitted an
%$N^{\rm th}$-order polynomial to the set of stars for which $t_{\rm
%B20,iso}/(+\sigma_{t,{\rm B20,iso}})\approx 3$, where
%$t_{\rm B20,iso}$ was the isochronal age reported by
%\citet{Berger_2020a_catalog}. We let the order of the fit vary from
%$N$=1 to 10, minimized the Bayesian Information Criterion,
%and found $N$=6.  For F and G dwarfs, this procedure failed to produce
%a reasonable locus because of shifts in the relative precision of
%the isochrone ages.  We therefore visually selected KOIs that appeared
%to be near the ``main sequence'', fitted a separate polynomial to
%their locations, and then stitched the resulting polynomials from the
%F and G dwarfs to that of the K dwarfs.}
the general aim is to flag stars for which there is evidence based on
their location in the $\log g$--$T_{\rm eff}$ plane that our
gyrochronology model may not be reliable.  
%The rotation period
%distribution of sources excluded in this manner includes a relatively
%larger fraction of ``rapid rotators'' than the main sequence sources,
%likely due to a combination of astrophysical and systematic effects.
See for instance Figure~8a of \citet{2022AJ....164..137K} or
Figures~12 and 13 of \citet{2023ApJS..268....4F} for examples of how
the period distribution of this population differs from stars near the
main sequence.

@@

In part because many of our EWs are upper limits, many of the lithium
ages are lower limits.  In such cases,  the inferred age posterior is
strongly influenced by the assumed intrinsic dispersion in
the lithium model, and by the prior.  In these instances, we
quote the 3$\sigma$ (99.7$^{\rm th}$ percentile) limits.

@@

{\it Occurrence Rates}:
The age-dependent trends predicted for exoplanet populations, such as
the Kelvin-Helmholtz cooling of mini-Neptunes \citep{Gupta_2019},
time-dependent carving of the photoevaporation desert
\citep{Owen_Lai_2018}, and the time evolution of the radius valley
\citep{Rogers_2021} can be explored using our data.  We defer this
analysis to a separate publication.
Some care is required since beyond age,
exoplanet demographics also depend on stellar metallicity and mass
\citep[e.g.][]{Petigura_2018}.

@@

\startlongtable
\begin{deluxetable*}{llllllllrrrrrrl}
  \tabletypesize{\scriptsize}
  \tablecaption{Ages of Kepler planets and planet candidates.  This
  version of the table is truncated to include only the youngest
  confirmed Kepler planets, sorted by the minimum of either the
  rotation or lithium-based age.  The full machine-readable table
  contains ages and age limits for \nnonfopkoissomeageinfo\ non-false
  positive KOIs with MES$>$10.  
  A bash script to decode any given $Q_{\rm star}$ quality flag is
  {\bf
  \href{https://gist.github.com/lgbouma/20368253f1a98da1b39cf32fdda0be13}{available
  online}}.  A python script to select stars with specific bit flags
  is {\bf \href{https://gist.github.com/lgbouma/87ad8bde42625e766a5a8857cc5a183a}{also available}}.
  \label{tab:planets}}
  %\toprule
  %\midrule
  %\endhead
  \startdata
  KOI & Kepler &  $T_{\rm eff}$ & $P_{\rm rot}$ & EW$_{\rm Li}$ &
  $t_{\rm gyro}$ & $t_{\rm Li} $ & Consistent? & $R_{\rm p}$ & $P_{\rm orb}$ &
  $Q_{\rm planet}$ & $Q_{\rm star}$ &  Spec? & Comment \\
  -- &   -- & K & days & m\AA & Myr & Myr &  str &   Earths &    days &       int  & int & bool & -- \\
  \hline
  \input{table_subgyr_confirmed} %\\
% \rule
%  \input{table_subgyr_candidate}
  \enddata
  \tablecomments{Two checkmarks denotes systems with two-sided $t_{\rm
  gyro}$ and $t_{\rm Li}$ for which no important quality flags are
  raised.  One checkmark denotes two-sided $t_{\rm gyro}$ only. 
  ``Confirmed'' planets appear in the machine-readable
  version before``candidate'' planets.  The bit quality flags for the
  rotation-based ages, $Q_{\rm star}$, are described in
  Section~\ref{subsec:flags}.  Concisely summarized, they are:
  Bit 0: $T_{\rm eff}/{\rm K} \in [ 3800-6200]$ ?
  Bit 1: $\log g$$<$4.2?
  Bit 2: $M_{\rm G}$<3.9 or $M_{\rm G}>8.5$?
  Bit 3: In KEBC?
  Bit 4: Large $d_{\rm xm,Kep-Gaia}$?
  Bit 5: Confused Kep-Gaia crossmatch?
  Bit 6: Gaia DR3 non-single star?
  Bit 7: RUWE$>$1.4?
  Bit 8: Crowded?
  Bit 9: Far from main sequence?
  Bit 10: \citetalias{Santos_2021} CP/CB?
  Bit 11: $P_{\rm rot}$ not in the homogeneous
  \citetalias{Santos_2019} \citetalias{Santos_2021} sample?
  As an example, Kepler-1627 has $Q_{\rm star}$ flagged with bit 10 and bit 7.
  The analogous planet quality bitmask, $Q_{\rm planet}$, has the
  following meaning.
  Bit 0: Candidate not reliable?
  Bit 1: MES$<$10?
  Bit 2: Grazing?
  For a star to have a high likelihood of being ``reliable for
  gyrochronology'' we suggest $Q_{\rm star}$ bits zero to nine to
  not be raised, and for a planet to be ``reliabe'', we suggested
  $Q_{\rm planet}$ to be zero.
  }
\end{deluxetable*}
